
\chapter{Basics}
\label{chap:basics}


XX nette einleitung


\section{Wall-bounded turbulence}\label{sec:turbulenceTheory}

% - boundary layer develops const in flow direction
% - with increasing boundary layer thickness
% - wall shear stress not know a priori
% - behaviour in inner layer ess. same as in channel flow
% - equation for pressure gradient
% %--> U_0 = const pressure gradient = 0
% - definition of boundary layer thickness, displ. thickness, momentum thickness
% - corresponding Reynolds numbers
% - shape factor

The description of fluids is based on three physical principles.
The first  states that a fluid can be treated as a continuum \citep{schlichting_boundary-layer_2017}.
In addition we take the viscous property of fluids into account which for a Newtonian fluid relates shear stresses to XX \citep{tennekes_first_1972}
And as a last assumption we want the mass conservation to hold.
With these assumptions we can define the \benennung{continuity equation} for viscous incompressible flows as

\begin{equation}
    \frac{\partial \rho}{\partial t} + \frac{\partial (\rho u_i)}{\partial x_i} = \frac{\partial u_i}{\partial x_i} = 0
\end{equation}

The second important equation describes the momentum conservation of a fluid.
This so-called \benennung{\acrfull{nse}} can be written as 

\begin{equation}
    \frac{\mathrm{D}u_i}{\mathrm{D}t} = \frac{\partial u_i}{\partial t} + u_j \frac{\partial u_i}{\partial x_j} = -\frac{1}{\rho} \frac{\partial p}{\partial x_i} + \nu \frac{\partial^2 u_i}{\partial x_j \partial x_j}
\end{equation}

With these equations the motions of a fluid is fully described, since it holds for most kind of flows (XXQUELLE).
However, one can split the behaviour of fluids into three groups with different characteristic properties.
These categories are defined as laminar, turbulent and transitional. 
If a flow acts in the laminar regime its streamlines do not cross and the flow is time-independent \citep{davidson_turbulence_2015}.
IN the turbulnet regime the behaviour of the flow is chaotic and time-dependent (XXQUELLLE).
It shows irregular time and locations dependent random fluctuations \citep{pope_turbulent_2015}.
Through this behaviour in turbulent flows much more mixing in the fluid can be observed (XXQUELLE).
To describe the behaviour of a flow in terms of laminar or turbulent behaviour the so-called \benennung{Reynolds number} is defined as 
\begin{equation}
    \mathrm{Re} = \frac{U L}{\nu}
\end{equation}
where U and L are reasonable velocity or length scales and $\nu$ the viscosity of the fluid.
Flows with low Reynolds numbers show laminar flow behaviour.
With increasing Reynolds number the fluid comes first in the transitional region between laminar and turbulent until it reaches a fully turbulent state.
At high Reynolds numbers one can observe a separation of scales with different behaviours \citep{pope_turbulent_2015}.
One can split the turbulent motions into large and small scale motions.
The large scales are mostly controlled by the outer constraints of the flow e.g. its geometry and the boundary conditions.
The behaviour of the small-scale motions however depends on the viscosity and the receiving energy from the large scales \citep{pope_turbulent_2015}, which makes their behaviour universal and independent of the specific flow case.
This universal behaviour can be measured with statistics over long time averaging.

The most turbulent flows are bounded by at least one solid wall (XXQUELLE).
One can split them into \benennung{external} and \benennung{internal} flows. 
Internal flows are flows that are limited by outer walls like pipes or channels.
External flows are flows which flow around a solid body like an airfoil or a wall.
A characteristic of external flows is that the flow far away from the body being flowed around is not affected by it (XXQUELLE).
The behaviour of the flow near the wall is in all these cases very similar because of the universal small scales \citep{pope_turbulent_2015}XX umformulieren.
In the following section we will derive some definitions to describe this near-wall behaviour.
Since this work focusses on boundary layers along a wall, the following chapter contains some definitions specified for this case.


\subsection{Turbulent boundary layer theory} \label{subsec:TBLTheory}

Boundary layers occur when fluids with non-zero viscosity stream over a wall that is not moving. 
In these setups the so called \benennung{no-slip} boundary condition impacts the flow behaviour significantly.
It states that the velocity components of the fluid in wall-normal and streamwise direction need to be zero directly at the wall since there is no relative movement between the wall and the fluid \citep{schlichting_boundary-layer_2017}. 
The velocity far away from the wall, the so called \benennung{free-stream velocity} $U_0$, is usually much higher than zero which implies the existence of a transition region where an acceleration of the fluid occurs (XXQUELLE).
This so-called \benennung{boundary layer} is placed close to the wall. 
Depending on the Reynolds number the thickness of the boundary layer varies in a way that the boundary layer decreases with increasing Reynolds number \cite{schlichting_boundary-layer_2017}. 
The flow in the boundary layer can behave laminar or turbulent, whereas in this thesis we will focus on \acrfull{tbl}.
The boundary layer grows in streamwise direction and can be described via multiple characteristic quantities. 
The introduction of these quantities requires a consistent coordinate system: 
We define $x$ as the streamwise direction with its corresponding velocity $u$, $y$ represents the wall-normal direction with velocity component $v$ and $z$ describes the spanwise direction with velocity component $w$.
In the free stream far away from the wall we assume $U_0$ being the dominant velocity component. 
For simplicity this free-stream velocity is often set to be 1. XX diesen satz vlt eher bei non-dim? XX
The streamwise velocity component inside the boundary layer is referred to as $u(x,y)$.
To describe the thickness of a boundary layer mainly three different parameters are used: 
the boundary-layer thickness $\delta_{99}(x)$, which describes the distance from the wall at which the velocity $u(x,y)$ derives $99\%$ of the free-stream velocity;
the displacement thickness $\delta^*$, which defines the displacement of an inviscid fluid to preserve the same mass flow;
and the momentum thickness $\theta$  which defines the displacement of an inviscid fluid to fulfil the conservation of momentum.
These quantities can be derived with the following equations
\begin{gather}\label{eq: tblThickness}
    u (x, \delta_{99}(x)) = 0.99 U_0 \\
    \delta^*(x) = \int_0^\infty \left( 1 - \frac{u(x,y)}{U_0} \right) \mathrm{d}y \hspace{0.8cm}
     \theta(x) = \int_0^\infty \frac{u(x,y)}{U_0} \left( 1 - \frac{u(x,y)}{U_0} \right) \mathrm{d}y
\end{gather}

In this thesis we use the momentum thickness $\theta$ to characterize the height of the boundary layer, since it can be measured more accurate than the boundary-layer thickness (XXQUELLE). 
To describe the boundary-layer with the Reynolds number, different Reynolds numbers based on the introduced displacement measures are defined as

\begin{gather}
    \mathrm{Re}_x = \frac{U_0 x}{\nu}, \quad
    \mathrm{Re}_{\delta_{99}} = \frac{U_0 \delta_{99}}{\nu}, \quad
    \mathrm{Re}_{\delta^*} = \frac{U_0 \delta^*}{\nu}, \quad
    \mathrm{Re}_\theta = \frac{U_0 \theta}{\nu}. \\
    \nu : \text{viscosity} \nonumber
\end{gather}


%XX das hier weglassen statdessen mehr über non-dim und C_p/ C_f? 
From the \acrfull{nse} we can derive the boundary layer equations for a 2D boundary layer over a fixed flat wall for an incompressible fluid. 

\begin{align}
    \frac{\partial u}{\partial x} + \frac{\partial v}{\partial y} &= 0 \\
    u \frac{\partial u}{\partial x} + v \frac{\partial u}{\partial y} &= -\frac{1}{\rho} \frac{\partial p}{\partial x} + \nu \frac{\partial^2 u}{\partial y^2} \\
    0 &= - \frac{1}{\rho} \frac{\partial p}{\partial y} \label{eq:NSEy}
\end{align}

XX notwendig?

With this equation we can define the wall shear stress $\tau_{w}$ as
\begin{equation}
    \tau_w := \mu \left. \frac{\mathrm{d}U}{\mathrm{d}y} \right|_w = -h \frac{\mathrm{d}p_w}{\mathrm{d}x}.
\end{equation}

X hier non-dim einführen
To study the phenomenons in a boundary layer it is required to define so called viscous scales since the ranges of dimensions and velocity variations are much smaller than in the free-stream.
Based in the wall shear stress the friction velocity $u_{\tau}$, the viscous length scale $\delta_{\nu}$ and the friction Reynolds number $Re_{\tau}$ are introduced as
\begin{gather}
    u_{\tau} = \sqrt{\frac{\tau_w}{\rho}} \hspace{1.0cm}
    \delta_{\nu} = \frac{\nu}{u_{\tau}} \hspace{1.0cm}
    Re_{\tau} = \frac{\delta_{99} u_\tau}{\nu}
\end{gather}

With these quantities the original length and velocity scales can be nondimensionalized in wall units as (fundamentals of tubr flows)
\begin{equation}
    y^+ = \frac{y}{\delta_v} = \frac{y u_\tau}{\nu}, \qquad
    u^+ = \frac{U}{u_\tau}.
\end{equation}

The advantage of these non-dimensional scales is that they are universal for all wall buonded flows what enables the comparison between flows with different physical dimensions. XX unbedingt QUELLE XX


XX hier noch loglaw 


\subsubsection{Reynolds stresses}
To measure velocities in a turbulent flow they are decomposed in a time averaged component $\langle \rangle$ and fluctuations $u_i'$. 

\begin{equation}
    u_i = \langle u_i \rangle + u_i' = U_i + u_i' \text{ with } \langle u_i'
\rangle = 0\end{equation}

represents the so-called \benennung{Reynolds decomposition}.
Teh application of the reynolds decomposition on the \acrshort{nse} leads to the \acrfull{rans}. 

\comment{
\begin{equation}
    \frac{\partial \langle u_i \rangle}{\partial t} + \frac{\partial \langle u_i \rangle \langle u_j \rangle}{\partial x_j} = -\frac{1}{\rho} \frac{\partial \langle p \rangle}{\partial x_i} + \frac{1}{\rho} \frac{\partial}{\partial x_j} \left(  \mu \left( \frac{\partial \langle u_i \rangle}{\partial x_j} + \frac{\partial \langle u_j \rangle}{\partial x_i} \right) + \langle u_i' u_j'\rangle \right)
\end{equation}
}

From this one can extract one term as the \benennung{Reynolds stresses}.
For turbulent flows the Reynolds stresses are important measures, since the transponrt of mean momentum is based on them \citep{moin_fundamentals_2024}. 
The tensor is defined as 
\begin{equation}
    R_{ij} =  \langle u_i' u_j'\rangle.
\end{equation}
The inner part $u_i' u_j'$ of the tensor elements represents the product of the velocity fluctuations.
To get the Reynolds stresses, the mean $\langle \rangle$ of this product is taken.
Although the mean of one fluctuation is zero, the product of two fluctuation components is not necessarily zero. 
The diagonal entries therefore, contain the normal stresses $\langle u' u'\rangle, \langle v' v'\rangle, \langle w' w'\rangle$.
The remaining entries contain the shear stresses.
Here the only non-zero entry is $\langle u' v'\rangle$ \citep{schlichting_boundary-layer_2017}.
The issue about this tensor is, that we cannot directly compute it.
With the continuity equation and the \acrshort{nse} in three directions we have four equations to derive the pressure and the three velocity components.
With the Reynolds stress tensor we have six additional terms we have to solve.
This problem is called the \benennung{closure problem} \citep{wilcox_turbulence_2006}.
To solve this, there are different approaches how to derive more equations such that all terms can be computed which is describe in detail in fundamental books like \citet{wilcox_turbulence_2006,pope_turbulent_2015} XXnoch eines.
One can derive the trasnport equations for the REynolds stresses by XXX \citep{mansour_reynolds-stress_1988}
For incompressible flows the non-dimensionalized transport equations can be stated as 
\begin{gather}
    0 = -\frac{\overline{\mathrm{D}}}{\overline{\mathrm{D}}t} \langle u_i u_j \rangle - T_{ij} + D_{ij} + \mathcal{P}_{ij} + \Pi_{ij} - \varepsilon_{ij}, \\[4pt]
    \begin{aligned}
        \mathcal{P}_{ij} &\colon \text{production rate,} & \quad \varepsilon_{ij} &\colon \text{dissipation rate,} \\
        T_{ij} &\colon \text{turbulent transport rate,} & \quad D_{ij} &\colon \text{viscous diffusion rate,} \\
        \Pi_{ij} &\colon \text{velocity pressure-gradient term,} & \quad \frac{\overline{\mathrm{D}}}{\overline{\mathrm{D}}t} \langle u_i u_j \rangle &\colon \text{convection}
    \end{aligned} \notag
\end{gather}
\comment{
\begin{equation}
\begin{aligned}
    0 = &-\left( \langle U \rangle \frac{\partial k}{\partial x} + \langle V \rangle \frac{\partial k}{\partial y} \right) + \mathcal{P} - \tilde{\varepsilon} + \nu \frac{\partial^2 k}{\partial y^2} \\
    &- \frac{\partial}{\partial y} \left\langle \frac{1}{2} v \boldsymbol{u} \cdot \boldsymbol{u} \right\rangle - \frac{1}{\rho} \frac{\partial}{\partial y} \langle v p' \rangle.
\end{aligned}
\end{equation}
\begin{equation}
    0 = -\frac{\overline{\mathrm{D}}}{\overline{\mathrm{D}}t} \langle u_i u_j \rangle - T_{ij} + D_{ij} + \mathcal{P}_{ij} + \Pi_{ij} - \varepsilon_{ij},
\end{equation}}


\subsection{Clauser parameter in turbulent boundary layers}

In boundary layers the free-stream (XX stimmt das?) pressure gradient plays a crucial role since it impacts the flow behaviour significantly (XXQUELLE).
From \autoref{eq:NSEy} we see that the pressure for a \acrshort{tbl} along a flat wall is constant in wall-normal direction.
In streamwise direction however it not necessarily has to be constant.
With Bernouli's equation the free-stream pressure is connected to the freestream velocity, such that the gradient in tstreamwise direction can be writte like
\begin{equation}\label{eq:bernoulli}
    -\frac{\mathrm{d}p_0}{\mathrm{d}x} = \rho\, U_0 \frac{\mathrm{d}U_0}{\mathrm{d}x}.
\end{equation}
This equation shows that an acceleration of the flow effects a negative pressure gradient, which is also called \benennung{\acrfull{fpg}}. 
In contrast, an increase of the velocity leads to a positive or \benennung{\acrfull{apg}}. 
\acrshort{tbl} with a constant freestream velocity therefore have no pressure gradient.
This \benennung{\acrfull{zpg}} configuration is well studied, since in plane \acrshort{tbl} it is the natural configuration (XXQUELLE).
However, it is possible to apply boundary conditions such that an adverse or favourable pressure gradient results.
To measure the pressure gradient a common used quantity is the so-called \benennung{Clauser parameter $\beta(s)$} which is defined as
\begin{equation}\label{eq:beta}
    \beta(s) = \frac{\delta^*(s)}{\tau_w(s)} \frac{\mathrm{d}p_w}{\mathrm{d}s},
\end{equation}

This non-dimensionalized parameter defines the ratio between the streamwise pressure force on the boundary layer and the wall shear stress \citep{clauser_turbulent_1954}.
In the literature this parameter is widely used to categorize the flow according to the introduced pressure regimes:

\begin{equation}
    \beta =
    \begin{cases}
        < 0, & \text{favourable pressure gradient (FPG)}, \\
        = 0, & \text{zero pressure gradient (ZPG)}, \\
        > 0, & \text{adverse pressure gradient (APG)}.
    \end{cases}
\end{equation}




\section{curved TBL}

Most research about \acrshort{tbl} focusses on \acrshort{tbl} across straight walls. (XXQUELLE) 
However, there are many real-world applications where the wall is not fully straight but shows curved sections as well (XXQUELLE). 
The effects in \acrshort{tbl} along curved walls are still poorly studied (XXQUELLE) since it covers some difficulties. 
Due to the curved shape a pressure gradient in streamwise direction builds up naturally which makes it hard to separate the source of observed physical effects \citep{so_experimental_nodate}. 
As \citet{so_experiment_1973} observed in their experiments without pressure adjustment the pressure decreases slow near the beginning of the curvature and then decreases fast once the curvature starts. 
Since these pressure variations have a high impact on the turbulent structures, an exclusive study of curvature effects requires a compensation of these pressure variations (XXQUELLE). 
Therefore, it is possible to study curved \acrshort{tbl} under zero-pressure gradient conditions (see XXX), or to impose a defined pressure gradient (see XXX). 
In experimental studies a defined pressure distribution is achieved by adjusting the position of the upper wall \citep{so_experimental_nodate, johnston_experimental_nodate, gillis_turbulent_1983,schwarz_convex_1996}. 
In simulations 
From these studies some general statements about the effects of curvature on the \acrshort{tbl} can be summarized.
First a boundary layer along a convex wall with \acrshort{zpg} shows smaller growth rate compared to a straight developing \acrfull{tbl} \citep{so_experimental_nodate}, \citep{meroney_turbulent_1975}. 
This phenomenon is caused by a pressure gradient in centripetal direction. 
It balances the centrifugal force of the fluid elements. Through the centrifugal force fluid particles with high velocities will be pushed to a new outer position. 
The pressure gradient forces the fluid particle in the opposite direction \citep{so_experimental_nodate}, \citep{meroney_turbulent_1975}.
\citet{gillis_turbulent_1983} observed a decrease in the turbulnet scales as consequence of a convex curvature. 
In addition, they found that the law of the wall coincides for the curved wall with the same parameters as for flat wall measurements. 
Besides experimental investigations, some numerical studies about the characteristics of curved \acrshort{tbl} are presented.

\comment{simulations}

Similar as in the experimental investigations a modification of the setup is necessary to compensate the streamwise pressure gradient \citet{appelbaum_systematic_2025}, XX. 
One possible approach to do so is the setup of a curved and a flat \acrshort{tbl} simulation and match their pressure distributions.
The naturally occuring pressure gradient in the curved domain is then imposed as boundary condition on the flat \acrshort{tbl} simulation.
With the flat plate as reference case one can thus isolate the effects of curvature.
This approach is used by \citet{spalart_direct_2024,pargal_adverse-pressure-gradient_2022}.
XXX hier bobke and schlatter einfügen
A similar approach \citet{bobke_history_2017} followed in their study about the effect of history on \acrshort{apg} \acrshort{tbl}.
To do this the $\beta(x)$ distribution from a wing \acrshort{dns} was copied on a flat \acrshort{tbl} through the imposition of corresponding velocity profiles along the top wall.
They studied various distributions and showed that even with matching $\beta(x)$ and $Re_{\tau}$ the geometry of the flow impacts the flow \citep{bobke_history_2017}.
Another possibilty of course is the study of curved setups without manipulation of the pressure as done by \citet{vinuesa_pressure-gradient_2017,tanarro_effect_2020}, nd XX.
\citet{vinuesa_pressure-gradient_2017} analyzed data from flows along a wing which were derived by \citet{hosseini_direct_2016}.
They studied the occuring pressure field and found that even small variations of $\beta$ impacted the development of the boundary layer compared to \acrshort{zpg} \acrshort{tbl} \citet{vinuesa_pressure-gradient_2017}.
\citet{tanarro_effect_2020} extended these studies and analyzed the different effects of local and integrated pressure gradient distributions.
However, these studies mainly focus on the effect of pressure on the flow characteristics refereeing to the argument that the pressure driven effects superpose the curvature effects \citep{patel_longitudinal_1997}.
\citet{yang_assessment_2016} studied a quite similar setup compared to the one used in the project to investigate curvature effects in \acrshort{tbl}. 
Their domain has a straight inlet, then a 60° circular bend section followed by a straight tail.
The \acrshort{rans} simulations they performed showed in comparison with \acrshort{dns} results, that accuracy of the tested turbulence models in insufficient in cases with large curvature \citep{yang_assessment_2016}.
In the study of \citep{appelbaum_systematic_2025} a concave \acrshort{tbl} was investigated. 
To achieve a \acrshort{zpg} distribution the pressure profile from the top wall of a plane \acrshort{tbl} was imposed on the top wall of the curved domain.
Since the pressure is the result of the flow behaviour it cannot be set directly an implicit approach was necessary to reach the desired pressure distribution \citep{appelbaum_systematic_2025}.
This overview of current research about curved \acrshort{tbl} shows that there is an interest in investigation of curvature effects on \acrshort{tbl}.
However, the separation of pressure and curvature effects is still an issue (XXQUELLE).
Many investigations focus on rebuilding the pressure distribution of a curved domain on a flat domain to get some understanding.
These studies already create some knowledge about curved \acrshort{tbl}.
However, the effects of curvature adn pressure gradient still interfere with each other.
The study of the isolated curvature in simulations is rare since the realization of the \acrshort{zpg} is not straight-forward.
The purpose of this project is to implement and test a new workflow for setting up a curved \acrshort{zpg} \acrshort{tbl}.
The used approach is outlined in \citet{schlatter_leveraging_nodate}.
To design a \acrshort{dns} domain with \acrshort{zpg} distribution the velocities along the upper wall will be manipulated.
The velocity componentn itself are extracted from optimized \acrshort{rans} simulaitons.
These \acrshort{rans} simulaitons are intergrated in an optmization algorithm where the domain shape is manipulated to derive a desired $\beta$ distribution \citet{schlatter_leveraging_nodate}.
The optimisation algorithm itself was introduced by \citet{morita_applying_2022}.
XX hier noch absatz 5.1 aus BO paper integrieren als motivation
The capability of the algorithm on curved domains was tested by \citet{ganagdhar_bayesian_2025}.
However, the results are only partly transferrable to the investigations from this study since the curvature studied by \citet{ganagdhar_bayesian_2025} had a much lower $\delta/R$ ratio. 
A detailed discussion about the transferability of other results is done in chapter XX.


\section{Simulation techniques} 
\label{sec:SimBasics}

For the simulation of turbulent flows differnet methods can be used. 
The decision about the used approach depends on the characteristics of the flow and the requirements for the solution.
One straight-forward approach is the \acrfull{dns}.
THis method requires solving of the \acrshort{nse} to compute $U(x,t)$.
\acrshort{dns} is the computational most expensive approach since all length and time scales are resolved \citep{pope_turbulent_2015} which one the other hand gives the most accurate results.
Especially if the small turbulent scales are of interest, the usage of \acrshort{dns} is inevitable.
Because of the high computational costs \acrshort{dns} are restricted to low to moderate Reynolds numbers and mostly used for research.
The simulation method of \acrshort{dns} is mainly used within this project.
Additional simulations were performed with \acrfull{rans} simulations.
In this approach the \acrshort{rans} are solved instead of the full \acrshort{nse}.
As a consequence only the mean velocity field is calculated.
The effect of turbulence is summarized in the Reynolds stress tensor.
To avoid the computation of this tensor multiple approaches were developed to model it.
Since the models have different characteristics and accuracies the model choice can have significant impact on the simulation (XX QUELLE).
These simulations are computational much easier to realize.
If the exact behaivour of the turbulent scales is not important but only is interested in the effect of the turbulnece on the flow \acrshort{rans} simulations build a reasonable compromise (XXQUELLE).
Therefore this simulation method finds application in multiple industrial environments.


\section{Optimization??}

\section{Tools}

XX das hier vlt in Basics?? XX

To work on this project multiple tools were used. Most of the fluid-flow simulations (korrekte bezeichnung??) were performed with DNS (see XX \autoref{sec:SimBasics}). To perform these simulations the spectral element code \name{neko} was used. Blabla über neko
When running simulaitons with \name{neko} it is possible to take statistics of the simulated flow during run time. Multiple settings about how the statistics should be recorded can be defined in the case file of the simulation. For example it is possible to define one or two axis along which the statistics should be averaged. These statistics are used to extract and compute certain measures of the flow. To do this some postprocessing of the recorded statistics is necessary. This is done in \name{python} where the package \name{pysemtools} is included. With this package data recorded in \name{neko} simulaitons can be extracted and interpolated from an SEM mesh into any arbitrary points (zu nah an github description). Then multiple statistical measures like the budget terms can be calculated. 
In addition to the DNS, some RANS simulations were performed as well. To run these simulations the open source software \name{OpenFOAM} is used. THis software is able to perform RANS simulations of fluid flows with diffenret turbulence models. To extract velocity fields etc from the simulation \name{Python} is used. 
All simulations are performed on the cluster of the RRZE of the FAU. Blabla über cluster
